Keep your Laziness in Check

### 2018-02-27 paper/main.tex
[paper] Editing

As a first example, consider purely functional queues, a very
popular data structure in Haskell. Programmers typically implement
such queues with two lists---a front list holding the queue elements
in FIFO order, and a back list holding the other elments in reversed
order. When the front list is empty, a dequeue operation requires a
reversal of the back list to refill the front list. These specific
dequeue operations take $O(n)$ time to finish, and are much slower
than their imperative counterparts. This occasional performance
degradation is compounded by persistence where the $O(n)$ operation
could be repeated many times, and can be unacceptable in real-time
systems.

However, Okasaki cleverly improved upon this design using lazy lists
that both incrementally performs the reversal of the back list, and
incrementally appends the reversal of the back list onto the front
list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
following \lk{rotate} function to incrementally reverse and append the
back list.
\begin{inlinecode}
rotate :: [a] -> [a] -> [a] -> [a]
rotate      []       []  as =                       as
rotate      []  (b : bs) as =     rotate [] bs (b : as)
rotate (f : fs)      []  as = f : rotate fs []      as
rotate (f : fs) (b : bs) as = f : rotate fs bs (b : as)
\end{inlinecode}
If we computed the expression \lk{rotate front back []}, its value
will be equivalent to \lk{front ++ (reverse back)}. However, the
function \lk{rotate} is much lazier than the naive reverse and append
operation. The function \lk{reverse} has to evaluate the entire
structure of the back list, which is not incremental. On the other
hand, the function \lk{rotate} produces a list immediately, and packs
the rest of the computation into the lazy list \lk{rest}.

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands (\Cref{fig:ListD}).
%
\begin{figure}[H]
\centering
\begin{inlinecode}
data ListD a = NilD
             | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
\caption{\lk{ListD} type}
\label{fig:ListD}
\end{figure}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and a nullary constructor \lk{T}
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor
from \lk{Thunk}. For example, given the list \lk{[1, 2, 3]}, the
value \lk{ConsD T (ConsD T T)} represents the \textit{demand} that
reveals the first 2 \lk{Cons} constructors of the list, but does not
evaluate any other parts of the list. To improve readability of demand
values, we will also use the syntax \lk{\_ : \_ : \_} where
appropriate.

@@

\lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example. \hzh{should we show what a
counter example looks like?}

>>>>

For our first example, we turn to a very popular data structure:
purely functional queues.  Such queues are typically implemented with
two lists---a front list holding the queue elements in FIFO order, and
a back list holding the other elments in reversed order. When the
front list is empty, a dequeue operation must reverse the back list to
refill the front list. These specific dequeue operations take $O(n)$
time to finish and are much slower than their imperative
counterparts, a performance degradation only compounded by
persistence---where the $O(n)$ operation could be repeated many times,
and unacceptable for many real-time systems.

Amongst his work on purely functional data structures\cite{Okasaki},
Okasaki cleverly improved upon this design using lazy lists that both
incrementally performs the reversal of the back list, and
incrementally appends the reversal of the back list onto the front
list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
following \lk{rotate} function to incrementally reverse and append the
back list.
\begin{inlinecode}
  rotate :: [a] -> [a] -> [a] -> [a]
  rotate      []       []  as =                       as
  rotate      []  (b : bs) as =     rotate [] bs (b : as)
  rotate (f : fs)      []  as = f : rotate fs []      as
  rotate (f : fs) (b : bs) as = f : rotate fs bs (b : as)
\end{inlinecode}
If we computed the expression \lk{rotate front back []}, its value
will be equivalent to \lk{front ++ (reverse back)}. However, unlike
the naive reverse-and-append operation, \lk{rotate} is much
lazier. A \lk{reverse} function has to evaluate the entire structure
of the back list, which is not incremental. On the other
hand, \lk{rotate} produces a list immediately, and packs the rest of
the computation into the lazy list \lk{rest}.

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands (\Cref{fig:ListD}).
\leo{Why is this a figure and not just inline? Especially if we're inlining
the figure!}
%
\begin{figure}[H]
\centering
\begin{inlinecode}
  data ListD a = NilD
               | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
\caption{\lk{ListD} type}
\label{fig:ListD}
\end{figure}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and \lk{T}, a nullary constructor,
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor. For example,
given the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)}
represents the \textit{demand} that reveals the first 2 \lk{Cons}
constructors of the list, but does not evaluate any other parts of the
list. To improve readability of demand values, we will also use the
syntax \lk{\_ : \_ : \_} where appropriate.

@@

\lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example. \hzh{should we show what a
counter example looks like?}\leo{Definitely!}
